The Numerical Accuracy of Computing Electrostatic Interaction in the Inhomogeneous and Correlated Molecular Systems: for the Ewald Summation, SPME and Staggered Mesh Ewald Methods.

### 2012-07-03 longRange.hetero/alltext/publish/resubmit/main.tex
fix typos, add fig.rho.eps

\newcommand{\recheck}[1]{{\color{red} #1}}
\renewcommand{\v}[1]{\textbf{\textit{#1}}}
\renewcommand{\d}[1]{\textsf{#1}}
% \allowdisplaybreaks

@@

\begin{abstract}
  In this work, we develop the accurate error estimates for
  three state-of-art algorithms of long-range electrostatic interaction
  in inhomogeneous and correlated molecular systems.
  They are the Ewald summation, the smooth particle mesh Ewald (SPME) and
  the staggered mesh Ewald methods. Two branches of force computation, namely
  the ik- and analytical differentiation, are considered. All the
  estimates are developed by proposing a more general framework:
  if the error force is of pairwise form,
  then the root mean square force error is composed by three additive parts,
  the homogeneity error, the inhomogeneity error and the correlation
  error. Computationally scalable estimates (estimating the errors
  at the cost $\mathcal O(N\log N)$) are developed for all the considered
  algorithms.
  The effectiveness of the proposed estimates and the important role of
  the correlation error are carefully checked and demonstrated by
  example systems.
\end{abstract}

@@

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a 6-dimensional parameter space.
The experiential choices may be either less accurate or wasting
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one satisfying a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid bilayer membrane
solving in water: the hydrophobic tail(s) of the lipids carry no charge, and
the charges in the system
are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  Several
research have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this research is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
framework for the pairwise interaction is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.
% These error estimates are only of theoretical interest, and
% not applicable, because the computational costs of the error estimates
% are $\mathcal O(N^2)$ (homogeneity and inhomogeneity error)
% and $\mathcal O(N^3)$ (correlation error).
% If the error force has a special form so that the error estimates
% can be written as convolutions, the computational costs
% are reduced to $\mathcal O(N\log N)$ and 
% $\mathcal O(N^2\log N)$, respectively.
The computational cost of the homogeneity and inhomogeneity
error is $\mathcal O(N\log N)$ when the error force kernel is
of convolution form.
The  correlation error can also be
estimated at a cost of $\mathcal O(N\log N)$
by the nearest neighbor approximation
technique, which assume the error  is dominated by  short-range correlation.
This
is the case for most the liquid and gas systems far from critical point.
Generally speaking,
when the error force can be written as a pairwise form, 
all estimate are derived easily under this framework.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME 
and the staggered method Ewald methods.   The estimates
of the homogeneity and inhomogeneity errors are tested by two
randomly distributed inhomogeneous charges systems,
% (i.e. the charges are not correlated),
and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

In this paper, we consider the system with periodic boundary condition.
Suppose it is
composed by $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.
We study the force of a testing particle located at $\v r$, which
feels interactions from all charges $q_1, q_2, \cdots, q_N$
in the system, and does not
exert force on them.
The \emph{error force} is defined by the difference between
the exact force and the calculated
force on the testing particle
\cite{wang2012}, and is denoted by $\Delta \v F(\v r)$.
The magnitude of the force computation error 
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r) = \sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means the ensemble average.  We also
estimate the the ensemble average of the error force (also called
\emph{mean error force}), namely $\langle\Delta\v F(\v r)\rangle$,
which plays an important role in the error analysis of the
inhomogeneous systems~\cite{wang2012}.

One of the main results of this research is the error estimate for

@@

  Let a periodic molecular system be composed by $N$ charged particles

@@

  Most of the error estimates assumed the independency of the

@@

\item Analog to the short-range error estimate~\cite{wang2012}, the

@@

\item A non-varnishing mean error force~\eqref{eqn:thm-meanf} was proved

@@

  most of the simulation cases, the locally neutral condition is

@@

O(N^2\log N)$, which is still not applicable.

@@

second part on the r.h.s. of the Eqn.~\eqref{eqn:split-rho-q2} is

@@

of neighbors in $\Omega_j$ is small comparing with the number of charges

@@

Therefore, the first part

@@

the number of neighbors is small comparing with the size of the system.

@@

is taken over all possible water directions 

@@

% For the second term, do variable change $\v r'' = \v r' + \v s$, we have:
% \begin{align}
%   \textrm{II} = & 
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s).
% \end{align}
% Split the integral region of the variable $\v s$ into the first
% neighbor shell $\mathcal B(\v r', R_s)$ and the reset $\mathbb R -
% \mathcal B(\v r', R_s)$. In the case of water, the first neighbor
% shell means the region where a single molecule occupies. Within this
% shell, the relative position of the oxygen and two hydrogens are
% always fixed. Outside of the first neighbor shell, the position of
% oxygens and hydrogens are belongs to different molecules, so the their
% positions are assumed to be independent. In most cases, the first
% neighbor shell is small comparing to the simulation box size,
% therefore,
% \begin{align}\nonumber
%   \textrm{II}_{\textrm{out}} = & \,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s) \\ \nonumber
%   = &\,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r') \rho_q(\v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_q(\v r' + \v s) \\
%   \approx &\,
%   q_i^2
%   [\v K\ast\rho_{q}]^2 (\v r_i)
% \end{align}

\subsection{Ewald summation and its error estimate}
The Ewald method divides the electrostatic interaction in to three
parts: the direct part, the reciprocal part and the correction
part:
\begin{equation}
E = E_{\textrm{dir}} + E_{\textrm{rec}} + E_{\textrm{corr}},
\end{equation}
where 
\begin {align}\label{eqn:tmp2}
E_{\textrm{dir}} & = \frac12 \sum^{\ast}_{\v n}
\sum_{i,j = 1}^{N} \frac{q_iq_j \textrm{erfc}(\beta \vert\v{r}_{ij} + \v{n}\vert)}
{\vert\v{r}_{ij} + \v{n}\vert} \\\label{eqn:tmp3}
E_{\textrm{rec}} & = \frac1{2\pi V} \sum_{\v m \neq 0}
\frac{\exp(-\pi^2\v m^2 / \beta^2)}{\v m^2} S(\v m) S(-\v m) \\\label{eqn:tmp4}
 E_{\textrm{corr}}& = -\frac\beta{\sqrt \pi} \sum_{i=1}^N q_i^2
\end {align}
The distance between the two particles is denoted by $\v r_{ij} = \v
r_i - \v r_j$.  The lattice in the real space is denoted
by $\v n = n_1 \v a_1 + n_2 \v a_2 + n_3 \v a_3$, where $\v a_\alpha,
\ \alpha=1,2,3$ are box vectors. the structure factor $S(\v m)$ is
defined by
\begin{equation}\label{sm1}
S(\v m) = \sum_{j=1}^N q_j \exp (2 \pi i \v m \cdot \v r_j),
\end{equation}
where $\v m = m_1 \v a_1^\ast + m_2 \v a_2^\ast + m_3 \v a_3^\ast$ is
the reciprocal space lattice vectors. $\v a_\alpha^\ast,\ \alpha = 1,
2, 3$ are conjugate reciprocal vectors of $\v a_\alpha$, defined by
$\v a_\alpha \cdot \v a_\beta^\ast = \delta_{\alpha\beta}$. and $V =
\v a_1 \cdot \v a_2 \times \v a_3$ is the volume of the box.

@@

As mentioned before, the 
optimal computational cost of the Ewald summation
is $\mathcal O(N^{3/2})$,
which becomes not applicable when the system
size is larger than several hundreds of charged particles.
Both the PME ans SPME methods
are designed to reduce
the computational cost  in nearly the same
way.
% Since the SPME method is more precise \cite{deserno1998mue1} and
% more popular than the PME, we will focus our discussion on the
% former. All the error estimates develop by the current paper can be
% easily extended to the PME method.
Starting from the reciprocal part of the Ewald summation,
they firstly  calculate the term $e^{2\pi i\v m\cdot\v r}$ on a uniform
mesh in the real space.  Therefore, all calculation of structure factors
can be
accelerated by the fast Fourier transform (FFT), that is why the
computational cost of PME/SPME is reduced to $\mathcal O(N\log N)$. Then,
for any particle position $\v r_i$, the value of $e^{2\pi i\v m\cdot\v
  r_i}$ is interpolated by the known values on the neighboring mesh
points.  The PME method uses Lagrangian interpolation, while the SPME
uses B-spline interpolation,  which is more precise, and has higher order
of smoothness.
The working parameters of these methods are
the splitting parameter $\beta$,
the cut-off radius in real space $r_c$,
the order of interpolation $n$ and
the number of mesh points on each direction:  $K_\alpha,\ \alpha = 1,2,3$.
Notice $K_\alpha/2$ is the truncation
of the reciprocal summation.

@@

The staggered mesh approach or originally call ``interlacing'' was
recently applied to SPME (staggered mesh Ewald~\cite{cerutti2009staggered}).
It calculates the
reciprocal force on two meshes,
one of which locates at the mesh element center
of the other, and then the two reciprocal forces are averaged.
It is shown that staggering the meshes
cancels error in the force computation, and greatly improves the
accuracy.
Implementing the staggered mesh method for SPME is
relatively simple, and involves only little modification to the
existing codes.
% : calculated the reciprocal force by the existing
% methods, either ik-differentiation or analytical differentiation, on
% the two staggered meshes, and then average these results.
% It is shown latter that the staggered mesh cencels
% high-wave-number terms in the error force, and 
% preserves the Newton's third law for the analytical differentiation.
Obviously, the staggered mesh ik-/analytical differentiation
cost twice as much on the
reciprocal part as the original version.

@@

From the precise calculation of the reciprocal summation to the SPME
fast algorithm, the electrostatic interaction is approximated by two
steps.  The first step is the truncation of the reciprocal summation,
i.e. Eqn. \eqref{eqn:tmp6}. The second step is the approximation of
$e^{2\pi i\v m\cdot\v r}$ by B-spline
interpolation. It has been shown that the
error due to the first step is negligible comparing with the second
step,
at least for the parameters of practical interest~\cite{wang2010optimizing}.
Numerical
examples supporting this argument in inhomogeneous systems are also
given in Section \ref{sec:tmp3}.  Therefore, to estimate the force
error, it is enough to compare the SPME force with the
\emph{truncated} Ewald force \eqref{eqn:tmp8}.

@@

The error force of the analytical differentiation is 
\begin{align}\nonumber
  \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)
  % =\;&
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % e^{2\pi i\v m\cdot \v r}
  % \sum_{j\neq i}q_j\,
  % A(\v m, -\v r_j)
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % [A(\v m, \v r)\,
  % e^{2\pi i\v m\cdot \v r} ]
  % \sum_{j\neq i}q_j\,
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\\nonumber
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % A(-\v m, \v r)\,e^{2\pi i\v m\cdot \v r} \nabla_{\v r}e^{2\pi i\v m\cdot\v r} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % e^{-2\pi i\v m\cdot \v r}
  % \nabla_{\v r} [\,A(\v m, \v r) \,e^{2\pi i\v m\cdot\v r}\,]  \\\nonumber
  =&\,
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\\\nonumber
  & +\,
  q
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,
  \v B(\v m, \v r)\,e^{2\pi i\v m\cdot \v r}
  \sum_{j}q_j\,
  e^{-2\pi i\v m\cdot\v r_j} \\\label{eqn:ana.error.force}
  & +\,
  q^2
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,\v B(\v m, \v r),
\end{align}
with $\v B(\v m, \v r)$ defined by
\begin{align}\nonumber
  \v B(\v m, \v r)
  &=
  \nabla_{\v r}\,A(\v m, \v r)\\\label{eqn:error-B}
  &=
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m)\,
  2\pi i l K_\alpha\v a_\alpha^\ast\,e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r}.
\end{align}
The last term in Eqn.~\eqref{eqn:ana.error.force} is due to the
self-interaction~\cite{cerutti2009staggered, ballenegger2011removal,
  neelov2010interlaced}, which does not depends on the coordinates of
particles. Therefore, it can be calculated explicitly, and subtracted
from the analytical force during the simulation. The computational cost of
the self-interaction term is low comparing with the SPME
force calculation. It is shown that the self-interaction  error
dominate when charge density is low~\cite{cerutti2009staggered}.
Therefore, we always remove the self-interaction term from
the analytical differentiation.
% , and call it as the original analytical
% differetiation when comparing it with the staggered mesh analytical
% differetiation.

@@

\begin{figure}
  \centering
  \includegraphics[]{rand1.rho.eps}
  \caption{
    The charge number density  of Example 2.
    The red line: positive charge. The blue line: negative charge.
    Since the charges are uniformly distributed
    on $y$ and $z$ direction, the densities are averaged on $y$ and $z$,
    and plotted as a
    function of $x$.
  }
  \label{fig:tmp-rho1}
\end{figure}

>>>>

\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}
\renewcommand{\v}[1]{\textbf{\textit{#1}}}
\renewcommand{\d}[1]{\textsf{#1}}
% \allowdisplaybreaks

@@

\begin{abstract}
  In this work, we develop the accurate error estimates for
  three state-of-art algorithms of long-range electrostatic interaction
  in inhomogeneous and correlated molecular systems.
  They are the Ewald summation, the smooth particle mesh Ewald (SPME) and
  the staggered mesh Ewald methods. Two branches of force computation, namely
  the ik- and analytical differentiation, are considered. All the
  estimates are developed by proposing a more general framework:
  if the error force is of pairwise form,
  then the root mean square force error is composed \redc{of} three additive parts,
  the homogeneity error, the inhomogeneity error and the correlation
  error. Computationally scalable estimates (estimating the errors
  at the cost $\mathcal O(N\log N)$) are developed for all the considered
  algorithms.
  The effectiveness of the proposed estimates and the important role of
  the correlation error are carefully checked and demonstrated by
  example systems.
\end{abstract}

@@

The aforementioned fast algorithms have achieved great success
over the past three decades.
However, the fact that the user should provide six working parameters
for these methods may be problematic: it is difficult to try out the
most efficient parameter set in such a 6-dimensional parameter space.
The experiential choices may be either less accurate or wasting
computational effort.
This problem can be solved by an
in-depth study of the error introduced by these algorithms, namely the
error estimate~\cite{kolafa1992cutoff, hummer1995numerical,
  petersen1995accuracy, deserno1998mue2, stern2008mesh,
  wang2010optimizing, neelov2010interlaced, ballenegger2012convert},
by which the error is described by a function of the working
parameters.  Using the error estimates, a work flow has been designed to
automatically determine the nearly optimal combination of parameters,
in the sense that it is the fastest one satisfying a demanding
accuracy~\cite{wang2010optimizing}. However, so far, all error
estimates assume the homogeneity of the charge distribution, and the
independency of any pair of charges (i.e. the charges are
uncorrelated), but neither of the assumptions is satisfactorily
fulfilled in most molecular
systems of practical interest.  For example, the lipid bilayer membrane
solving in water: the hydrophobic tail(s) of the lipids carry no charge, and
the charges in the system
are correlated by the covalent bonds, the van der Waals
interaction, the hydrogen bonding network, etc..  Therefore, the current
error estimates may be problematic, or even not applicable.  Several
papers have shown that due to the correlation of charges, the error estimates
obviously deviate from the real
error~\cite{deserno1998mue2,wang2010optimizing,ballenegger2012convert}.

The purpose of this manuscript is to develop a reliable  error
estimate in inhomogeneous and correlated molecular systems.
We estimate the force error because the application is the molecular
dynamics (MD) simulation.  Firstly, a general error estimate
framework for the pairwise interaction is set up. We prove that the
error is decomposed into three additive parts: the homogeneity error,
the inhomogeneity error and the correlation error~\cite{wang2012}.
% These error estimates are only of theoretical interest, and
% not applicable, because the computational costs of the error estimates
% are $\mathcal O(N^2)$ (homogeneity and inhomogeneity error)
% and $\mathcal O(N^3)$ (correlation error).
% If the error force has a special form so that the error estimates
% can be written as convolutions, the computational costs
% are reduced to $\mathcal O(N\log N)$ and 
% $\mathcal O(N^2\log N)$, respectively.
The computational cost of the homogeneity and inhomogeneity
error is $\mathcal O(N\log N)$ when the error force kernel is
of convolution form.
The  correlation error can also be
estimated at a cost of $\mathcal O(N\log N)$
by the nearest neighbor approximation
technique, which \redc{assumes} the error  is dominated by  short-range correlation.
This
is the case for most the liquid and gas systems far from critical point.
Generally speaking,
when the error force can be written as a pairwise form, 
all estimate are derived easily under this framework.
Following this framework, we
develop the error estimates for the Ewald summation, the SPME 
and the staggered method Ewald methods.   The estimates
of the homogeneity and inhomogeneity errors are tested by two
randomly distributed inhomogeneous charges systems,
% (i.e. the charges are not correlated),
and is proved to
be very sharp.
The error estimates are also tested in a real inhomogeneous water system,
in which the charge correlation plays a very important role
on the force error. By 
including the nearest neighbor approximation of
\redc{the} correlation error, the quality of the estimates
is impressively enhanced.
This paper is ended by conclusions and remarks in
section~\ref{sec:conclusions}.

@@

In this paper, we consider a system with periodic boundary condition.
Suppose it is
composed \redc{of} $N$ charged particles located at $\v r_1, \v r_2, \cdots,
\v r_N$ with charges $q_1, q_2, \cdots, q_N$, respectively.
We study the force \redc{on} a testing particle located at $\v r$, which
feels interactions from all charges $q_1, q_2, \cdots, q_N$
in the system, and does not
exert force on them.
The \emph{error force} is defined by the difference between
the exact force and the calculated
force on the testing particle
\cite{wang2012}, and is denoted by $\Delta \v F(\v r)$.
The magnitude of the force computation error 
is defined by the \emph{root mean square (RMS)} \emph{error}, which is
the square root of the second moment of the error force: $\mathcal
E(\v r) = \sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$.  The
notation $\langle\cdot\rangle$ means the ensemble average.  We also
estimate the the ensemble average of the error force (also called
\emph{mean error force}), namely $\langle\Delta\v F(\v r)\rangle$,
which plays an important role in the error analysis of the
inhomogeneous systems~\cite{wang2012}.

One of the main results of this research is the \redc{following} estimate for

@@

  Let a periodic molecular system be composed of $N$ charged particles

@@

  Most \redc{previous} error estimates assumed the independency of the

@@

\item \redc{Analogous} to the short-range error estimate~\cite{wang2012}, the

@@

\item A non-vanishing mean error force~\eqref{eqn:thm-meanf} was proved

@@

  most of the simulation cases, the \redc{local neutrality} condition is

@@

O(N^2\log N)$, which is still not \redc{tractable}.

@@

second \redc{term} on the r.h.s. of the Eqn.~\eqref{eqn:split-rho-q2} is

@@

of neighbors in $\Omega_j$ is small \redc{compared} with the number of charges

@@

Therefore, the first \redc{term}

@@

the number of neighbors is small \redc{compared} with the size of the system.

@@

\redc{are} taken over all possible water directions 

@@

% For the second term, do variable change $\v r'' = \v r' + \v s$, we have:
% \begin{align}
%   \textrm{II} = & 
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s).
% \end{align}
% Split the integral region of the variable $\v s$ into the first
% neighbor shell $\mathcal B(\v r', R_s)$ and the reset $\mathbb R -
% \mathcal B(\v r', R_s)$. In the case of water, the first neighbor
% shell means the region where a single molecule occupies. Within this
% shell, the relative position of the oxygen and two hydrogens are
% always fixed. Outside of the first neighbor shell, the position of
% oxygens and hydrogens are belongs to different molecules, so the their
% positions are assumed to be independent. In most cases, the first
% neighbor shell is small compared to the simulation box size,
% therefore,
% \begin{align}\nonumber
%   \textrm{II}_{\textrm{out}} = & \,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_{q^2}(\v r', \v r' + \v s) \\ \nonumber
%   = &\,
%   q_i^2
%   \int_{\mathbb R^3}\d d\v r'\v K(\v r_i - \v r') \rho_q(\v r')
%   \int_{\mathbb R^3 - \mathcal B(\v r', s)}\d d\v s\,
%   \v K(\v r_i - \v r' - \v s)\,
%   \rho_q(\v r' + \v s) \\
%   \approx &\,
%   q_i^2
%   [\v K\ast\rho_{q}]^2 (\v r_i)
% \end{align}

\subsection{Ewald summation and its error estimate}
The Ewald method divides the electrostatic interaction in to three
parts: the direct part, the reciprocal part and the correction
part:
\begin{equation}
E = E_{\textrm{dir}} + E_{\textrm{rec}} + E_{\textrm{corr}},
\end{equation}
where 
\begin {align}\label{eqn:tmp2}
E_{\textrm{dir}} & = \frac12 \sum^{\ast}_{\v n}
\sum_{i,j = 1}^{N} \frac{q_iq_j \textrm{erfc}(\beta \vert\v{r}_{ij} + \v{n}\vert)}
{\vert\v{r}_{ij} + \v{n}\vert} \\\label{eqn:tmp3}
E_{\textrm{rec}} & = \frac1{2\pi V} \sum_{\v m \neq 0}
\frac{\exp(-\pi^2\v m^2 / \beta^2)}{\v m^2} S(\v m) S(-\v m) \\\label{eqn:tmp4}
 E_{\textrm{corr}}& = -\frac\beta{\sqrt \pi} \sum_{i=1}^N q_i^2
\end {align}
The distance between the two particles is denoted by $\v r_{ij} = \v
r_i - \v r_j$.  The lattice in the real space is denoted
by $\v n = n_1 \v a_1 + n_2 \v a_2 + n_3 \v a_3$, where $\v a_\alpha,
\ \alpha=1,2,3$ are box vectors. the structure factor $S(\v m)$ is
defined by
\begin{equation}\label{sm1}
S(\v m) = \sum_{j=1}^N q_j \exp (2 \pi i \v m \cdot \v r_j),
\end{equation}
where $\v m = m_1 \v a_1^\ast + m_2 \v a_2^\ast + m_3 \v a_3^\ast$ is
the reciprocal space lattice \redc{vector}. $\v a_\alpha^\ast,\ \alpha = 1,
2, 3$ are conjugate reciprocal vectors of $\v a_\alpha$, defined by
$\v a_\alpha \cdot \v a_\beta^\ast = \delta_{\alpha\beta}$. and $V =
\v a_1 \cdot \v a_2 \times \v a_3$ is the volume of the box.

@@

As mentioned before, the 
optimal computational cost of the Ewald summation
is $\mathcal O(N^{3/2})$,
which becomes not \redc{tractable} when the system
size is larger than several hundreds of charged particles.
Both the PME \redc{and} SPME methods
are designed to reduce
the computational cost  in nearly the same
way.
% Since the SPME method is more precise \cite{deserno1998mue1} and
% more popular than the PME, we will focus our discussion on the
% former. All the error estimates develop by the current paper can be
% easily extended to the PME method.
Starting from the reciprocal part of the Ewald summation,
they firstly  calculate the term $e^{2\pi i\v m\cdot\v r}$ on a uniform
mesh in the real space.  Therefore, all calculation of structure factors
can be
accelerated by the fast Fourier transform (FFT), that is why the
computational cost of PME/SPME is reduced to $\mathcal O(N\log N)$. Then,
for any particle position $\v r_i$, the value of $e^{2\pi i\v m\cdot\v
  r_i}$ is interpolated by the known values on the neighboring mesh
points.  The PME method uses Lagrangian interpolation, while the SPME
uses B-spline interpolation,  which is more precise, and has higher order
of smoothness.
The working parameters of these methods are
the splitting parameter $\beta$,
the cut-off radius in real space $r_c$,
the order of interpolation $n$ and
the number of mesh points on each direction:  $K_\alpha,\ \alpha = 1,2,3$.
Notice $K_\alpha/2$ is the truncation
of the reciprocal summation.

@@

The staggered mesh approach or originally call ``interlacing'' was
recently applied to SPME (staggered mesh Ewald~\cite{cerutti2009staggered}).
It calculates the
reciprocal force on two meshes,
one of which locates at the mesh element center
of the other, and then the two reciprocal forces are averaged.
It is shown that staggering the meshes
cancels error in the force computation, and greatly improves the
accuracy.
Implementing the staggered mesh method for SPME is
relatively simple, and involves only \redc{a small} modification to the
existing codes.
% : calculated the reciprocal force by the existing
% methods, either ik-differentiation or analytical differentiation, on
% the two staggered meshes, and then average these results.
% It is shown latter that the staggered mesh cencels
% high-wave-number terms in the error force, and 
% preserves the Newton's third law for the analytical differentiation.
Obviously, the staggered mesh ik-/analytical differentiation
cost twice as much on the
reciprocal part as the original version.

@@

From the precise calculation of the reciprocal summation to the SPME
fast algorithm, the electrostatic interaction is approximated by two
steps.  The first step is the truncation of the reciprocal summation,
i.e. Eqn. \eqref{eqn:tmp6}. The second step is the approximation of
$e^{2\pi i\v m\cdot\v r}$ by B-spline
interpolation. It has been shown that the
error due to the first step is negligible \redc{compared} with the second
step,
at least for the parameters of practical interest~\cite{wang2010optimizing}.
Numerical
examples supporting this argument in inhomogeneous systems are also
given in Section \ref{sec:tmp3}.  Therefore, to estimate the force
error, it is enough to compare the SPME force with the
\emph{truncated} Ewald force \eqref{eqn:tmp8}.

@@

The error force of the analytical differentiation is 
\begin{align}\nonumber
  \Delta\v F_{\textrm{rec}}^{\textrm{ana}}(\v r)
  % =\;&
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % e^{2\pi i\v m\cdot \v r}
  % \sum_{j\neq i}q_j\,
  % A(\v m, -\v r_j)
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % \nabla_{\v r}
  % [A(\v m, \v r)\,
  % e^{2\pi i\v m\cdot \v r} ]
  % \sum_{j\neq i}q_j\,
  % e^{-2\pi i\v m\cdot\v r_j} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\\nonumber
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % A(-\v m, \v r)\,e^{2\pi i\v m\cdot \v r} \nabla_{\v r}e^{2\pi i\v m\cdot\v r} \\\nonumber
  % & +\,
  % q^2
  % \sum_{
  %   \begin{subarray}{c}
  %     \vert m_{\alpha}\vert < K_\alpha/2\\
  %     \v m\neq 0
  %   \end{subarray}}
  % -2f(\v m)\,
  % e^{-2\pi i\v m\cdot \v r}
  % \nabla_{\v r} [\,A(\v m, \v r) \,e^{2\pi i\v m\cdot\v r}\,]  \\\nonumber
  =&\,
  \Delta\v F_{\textrm{rec}}^{\textrm{ik}}(\v r)\\\nonumber
  & +\,
  q
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,
  \v B(\v m, \v r)\,e^{2\pi i\v m\cdot \v r}
  \sum_{j}q_j\,
  e^{-2\pi i\v m\cdot\v r_j} \\\label{eqn:ana.error.force}
  & +\,
  q^2
  \sum_{
    \begin{subarray}{c}
      \vert m_{\alpha}\vert < K_\alpha/2\\
      \v m\neq 0
    \end{subarray}}
  -2f(\v m)\,\v B(\v m, \v r),
\end{align}
with $\v B(\v m, \v r)$ defined by
\begin{align}\nonumber
  \v B(\v m, \v r)
  &=
  \nabla_{\v r}\,A(\v m, \v r)\\\label{eqn:error-B}
  &=
  \sum_{\alpha=1}^3\sum_{l\neq 0}
  Z_{\alpha,l}(\v m)\,
  2\pi i l K_\alpha\v a_\alpha^\ast\,e^{2\pi i l K_\alpha\v a^\ast_\alpha\cdot \v r}.
\end{align}
The last term in Eqn.~\eqref{eqn:ana.error.force} is due to the
self-interaction~\cite{cerutti2009staggered, ballenegger2011removal,
  neelov2010interlaced}, which does not depends on the coordinates of
particles. Therefore, it can be calculated explicitly, and subtracted
from the analytical force during the simulation. The computational cost of
the self-interaction term is low \redc{compared} with the SPME
force calculation. It \redc{has been} shown that the self-interaction  error
\redc{dominates} when the charge density is low~\cite{cerutti2009staggered}.
Therefore, we always remove the self-interaction term from
the analytical differentiation.
% , and call it as the original analytical
% differetiation when comparing it with the staggered mesh analytical
% differetiation.

@@

\begin{figure}
  \centering
  \includegraphics[]{rand1.rho.eps}
  \caption{
    The charge number density  of Example 1.
    The red line: positive charge. The blue line: negative charge.
    Since the charges are uniformly distributed
    on $y$ and $z$ direction, the densities are averaged on $y$ and $z$,
    and plotted as a
    function of $x$.
  }
  \label{fig:tmp-rho1}
\end{figure}
